Temperature and moisture effects on greenhouse gas emissions from deep active-layer boreal soils

### 2016-04-27 manuscript/cpcrw_manuscript.Rmd
Incorporated VLB and APS comments
Text clarified in a number of spots; discussion beefed up (though a
number of areas need to be beefed up still); figure 4 now has Tukey’s
HSD labels.

---
title: Temperature and moisture effects on greenhouse gas emissions from deep active-layer boreal soils
author: "Ben Bond-Lamberty^1^, A. Peyton Smith^2^, Vanessa Bailey^2^"
output: html_document
---
^1^Joint Global Change Research Institute, DOE Pacific Northwest National Laboratory, College Park, MD USA; 
bondlamberty@pnnl.gov

@@

@@

Rapid climatic changes, rising air temperatures, and increased fires are expected to drive permafrost degradation and soil carbon (C) cycle changes in many high-latitude ecosystems. How these soils will respond to changes in their temperature, moisture, and overlying vegetation is highly uncertain, but critical to understand given the large soil C stocks in these regions. We used a laboratory experiment to examine how temperature and moisture control CO~2~ and CH~4~ emissions from mineral soils sampled from the bottom of the annual active layer, i.e. directly above permafrost, in an Alaskan boreal forest. Gas emissions from thirty cores, subjected to two temperatures and either field moisture conditions or experimental drought, were tracked over a 100-day incubation; we also measured a variety of physical and chemical characteristics of the cores. Gravimetric water content was `r pn(gwcbegin$mean, 2)` ± `r pn(gwcbegin$sd, 2)` (unitless) at the beginning of the incubation; cores kept at field moisture were unchanged at the end, but drought cores had declined to `r pn(gwcendmean["Drought"], 2)` ±  `r pn(gwcendsd["Drought"], 2)`. Carbon dioxide fluxes were strongly influenced by incubation chamber temperature, core water content, and percent soil N. Methane fluxes were most strongly correlated with percent N, but neither temperature nor water content was a significant first-order predictor of CH~4~ fluxes, however. The cumulative production of C from CO~2~ was over `r minmag` order of magnitudes higher than that from CH~4~. The CO~2~ temperature sensitivity (i.e. Q~10~) was `r pn(Q10_cumulative["Field moisture", "CO2"], 1)` and `r pn(Q10_cumulative["Drought", "CO2"], 1)` for the field moisture and drought treatments, respectively. Deep but unfrozen high-latitude soils have been shown to be strongly affected by long-term experimental warming, and these results help us understand the potential future dynamics of such soils with future climate and permafrost changes.

#### Introduction

High latitude ecosystems are being subjected to rapid changes in climate {IPCC, 2013 #3916} and fire occurrence and intensity {Kasischke, 2010 #2631}, notably in northwestern North America and Alaska {Hinzman, 2005 #4315}{Ju, 2016 #4320}. This will have a wide variety of effects on both ecosystems and society: in particular, rising temperatures and increasing fire will likely result in permafrost degradation {Pastick, 2015 #4311}{Zhang, 2015 #4372} and changes in soil temperature, with subsequent hydrology changes that will influence soil greenhouse gas (GHG) fluxes to the atmosphere. Such fluxes are a large component of the global C cycle and, because of the high C stocks of northern soils {Tarnocai, 2009 #2231}, could result in a significant and positive climate feedback {Treat, 2015 #4217}{Koven, 2011 #3092}.

The magnitude, timing, and form-in particular as methane (CH~4~) or carbon dioxide (CO~2~)-of such any such feedback remain highly uncertain {Schuur, 2015 #4319}. While northern soils hold enormous quantities {Tarnocai, 2009 #2231} of potentially mineralizable soil organic carbon (SOC), vegetation succession dynamics promote permafrost resilience to even large temperature changes {Jorgenson, 2010 #2627}. Such dynamics may however be disrupted by increased fire disturbance {Johnstone, 2010 #2465}. In addition, the stability of SOC is itself highly uncertain, as it depends on soil temperature and moisture, the ages of and ratio between the carbon (C) and nitrogen (N) pools {Weiss, 2015 #4317}, and its protection from competent microorganisms, enzymes, and resources {Bailey, 2012 #3624}, whether by organomineral sorption, chemical lability, or physical location {Schmidt, 2011 #3135}.

Temperature and moisture typically have strong and often interactive influences on soil GHG emissions. Laboratory incubations, field observations, and meta-analyses have documented increased CO~2~, and under some conditions CH~4~ {Olefeldt, 2013 #3456}, fluxes with rising temperature {Davidson, 2006 #1669}{Hashimoto, 2015 #4231}{Treat, 2015 #4217}. GHG responses to wetting and thawing dynamics are much less certain, with substantial variability between studies {Kim, 2012 #2863}. The anaerobic conditions common following permafrost thaw are expected to lower CO~2~ emissions but increase those of CH~4~ {Treat, 2015 #4217;Treat, 2014 #4213}, and such interactions are critical to examine in the course of long-term incubation experiments {Elberling, 2013 #3720}. More uniformly across the landscape, however, decadal warming, and perhaps drying, trends in Alaska {Bieniek, 2014 #4389} will work to counteract these effects, with uncertain consequences for the carbon cycle and climate system.

The goal of this study was to examine how temperature and moisture control C loss and GHG (CO~2~ and CH~4~) emissions from soils sampled from the bottom of the annual active layer–i.e., directly above permafrost–in an Alaskan boreal forest. Most previous studies have focused on surface soils or permafrost soils, neglecting deep active-layer soils that were identified as subject to strong effects from a two-decade warming experiment in the Alaskan Arctic {Sistla, 2013 #3666}. We also aimed to characterize the chemical and biological properties of these soils as they were subjected to a 100-day incubation at different temperatures, subjecting some cores to drying treatments. We hypothesized that (i) CO~2~ would be the dominant pathway for C loss in these largely aerobic soils; (ii) soils maintained at field moisture and high temperature would lose more C-CO~2~ than cores incubated at 4˚C, due to increased aerobic and anaerobic microbial activity; and (iii) core CH~4~ fluxes would be sensitive only to temperature, as no anaerobic conditions were planned for the cores.

#### Methods

**Field sampling**

The field component of this research took place in Caribou-Poker Creeks Research Watershed (CPCRW), part of the Bonanza Creek LTER (http://www.lter.uaf.edu/research/study-sites-cpcrw). CPCRW is located in the Yukon-Tanana Uplands northeast of Fairbanks, AK, a part of the boreal forest that has seen strong increases in air temperature and forest browning {Ju, 2016 #4320} over several decades. Annual average air temperature was -2.5 °C, and annual average precipitation 400 mm {Petrone, 2006 #4313}. The watershed's lowlands and north-facing slopes are dominated by black spruce (*Picea mariana*), feathermoss (*Pleurozium schreberi* and others), and *Sphagnum* spp.; the drier south slopes tend to be deciduous with a mixture of quaking aspen (*Populus tremuloides*), paper birch (*Betula neoalaskana*), and patches of alder (*Alnus crispa*).

@@

**Laboratory incubation**

In the lab, the soil cores were stored at 4 °C for several days until they were weighed and prepared for incubation. At that point (11-12 August 2015), three fragmented or otherwise damaged cores were discarded, and the remaining cores were randomly assigned to one of six groups (N=6 in each group). These included two incubation temperatures (4 and 20 °C), each with two moisture treatments: one in which soil moisture was maintained at field conditions (~28% moisture by volume), and a drought treatment in which no water was added and cores were allowed to dry down to ~5% moisture by volume. The fifth group was a 20 °C "controlled drought" one, in which water was added so that these cores' moisture status would close match those of the 4 °C "drought" cores, which we anticipated would dry more slowly than their 20 °C counterparts. The final 6-core group was used for destructive, pre-incubation measurements including moisture content, pH, soil carbon and nitrogen, and bulk density. Subsamples were collected and stored at -20 °C for dissolved organic carbon measurements.

On 18 August 2015 cores were placed into one of two growth chambers (Conviron CMP6050 Control Systems, Winnipeg, Canada) maintained at 4 and 20 °C temperatures and 70% relative humidity and allowed to equilibrate for two weeks. Starting on 31 August 2015 we measured the cores' mass and GHG (CH~4~ and CO~2~) evolution four times in the first week, then twice per week for the first month, and then once per week for the rest of the 100-day incubation. Throughout the incubation, cores were mounted on porous ceramic plates (Soil Moisture Equipment Corp., Santa Barbara, CA, USA) so that, when the plates were placed in contact with water, water would move up into the cores via capillary action. The "drought" cores were mounted on dry plates, but not allowed to drop below 5% water content. When necessary, cores received additional wetting from the top to maintain their water status at the desired level.

@@

The incubation experiment concluded on 9 December 2015, following the final CO~2~ and CH~4~ readings. Each soil core was maintained at the treatment-dependent temperature and moisture content (by mass) until removed for destructive sampling, December 14-18, 2015. Sub-samples were collected and composited throughout each soil core for dissolved organic carbon analysis (110 ± 24 g dry mass equivalent) and dry-mass calculations (~28 g each). The remaining core material was air-dried and separated into particles (>2 mm diameter) and soil (≤2 mm) using a U.S. Standard Test Sieve No. 10 (Fisherbrand,  Pittsburg, PA, USA). The dry mass and volume of soil were used in calculations of gravimetric and volumetric soil moisture content, respectively {Gardner, 1986 #4345}. Soil volume was caclulated as the total core volume minus the volume of particles >2 mm diameter, with the latter determined by water displacement. Air-dried soil and sub-samples stored at -20 °C were sent to the Agricultural and Environmental Services Laboratory at the University of Georgia Extension in February 2016 for total C, N, and dissolved organic C. Samples were combusted in an oxygen atmosphere at 1350 °C, and measured for gaseous C and N using a Elementar Vario Max CNS. DOC was measured using a Shimadzu 5000 TOC Analyzer.

**Data and statistical analysis**

@@

Anomalous data were excluded based on their gas fluxes being more than `r CO2_EXCLUDE_DEVS` (for CO~2~) or `r CH4_EXCLUDE_DEVS` (for CH~4~) mean absolute deviations {Davies, 1993 #3138} from the treatment mean within a 10-day period, for a given treatment and temperature. We excluded `r mad_outlier_count` of `r pre_outlier_count` (`r pn(mad_outlier_count/pre_outlier_count*100, 1)`%) measurements for this reason. If the coefficient of variability of fluxes from any core on a single day exceeded `r pn(MAX_CO2_CV*100, 0)`%, the entire core was also excluded for that day (`r cv_outlier_count` data points, `r pn(cv_outlier_count/pre_outlier_count*100, 1)`%). Other data (`r pn(rn/sdn*100, 1)`%) were removed because of known instrument problems, e.g. the analyzer was left running after leaving a chamber. The final number of valid flux samples from the 100-day incubation was `r sum(!fluxdata$outlier)`.

@@

The effects of temperature, gravimetric water content, C percent, N percent, and DOC concentration on instantaneous gas fluxes were evaluated using a linear mixed-effects model fit by the R function *lme* in the R 'nlme' package, version `r packageVersion("nlme")`. Because the dependent variable (CO~2~ or CH~4~ flux) was non-normally distributed, it was transformed using a natural-logarithm (+`r FLUX_ADDITION` µg C g C^-1^ day^-1^ to ensure all positive fluxes, following Treat et al. 2015) transformation. Core number was treated as a random effect in the model. We then used performed stepwise model selection by Akaike's information criterion (AIC) using the *stepAIC* function in the R 'MASS' package, version `r packageVersion("MASS")`. A linear mixed-effects model was also used to evaluate the effect of treatment on core water content.

@@

#### Results

```{r coredata, echo=FALSE}
bd <- with(coredata, SoilDryMass_g / SoilVolume_cm3)
schist <- with(coredata, NonsoilMassLarge_g / (NonsoilMassLarge_g + SoilDryMass_g)) * 100
sdm <- coredata$SoilDryMass_g
```

The `r nrow(coredata)` experimental cores had a bulk density of `r pn(mean(bd), 2)` ± `r pn(sd(bd), 2)` (mean ± sd) g cm^-3^. Large (>2 mm) particles, primarily schist, comprised `r round(mean(schist), 0)`% ± `r round(sd(schist), 0)`% of the cores' total mass. Soil (≤2 mm) dry mass was `r round(mean(sdm), 0)` ± `r round(sd(sdm), 0)` g. Sample DOC was `r pn(cndata_summary["DOC_mg_kg_mean"], 2)` ± `r pn(cndata_summary["DOC_mg_kg_sd"], 2)` mg kg^-1^. Carbon content was `r pn(cndata_summary["C_percent_mean"], 2)`% ± `r pn(cndata_summary["C_percent_sd"], 2)`%, while N content was `r pn(cndata_summary["N_percent_mean"], 2)`% ± `r pn(cndata_summary["N_percent_sd"], 2)`%. Neither temperature was treatment exerted any significant effect on these properties (P > 0.1 for all).

Gravimetric water content was `r pn(gwcbegin$mean, 2)` ± `r pn(gwcbegin$sd, 2)` (min `r pn(gwcbegin$min, 2)`, max `r pn(gwcbegin$max, 2)`) at the beginning of the incubation (**Figure 1**). "Field moisture" cores were on average unchanged (`r pn(gwcendmean["Field moisture"], 2)` ± `r pn(gwcendsd["Field moisture"], 2)`) at the end of the incubation, but both the drought treatments, which did not differ from each other in their effect on gravimetric water content (P = `r pn(p_trt_wc, 3)`), had declined to `r pn(gwcendmean["Drought"], 2)` ±  `r pn(gwcendsd["Drought"], 2)`. Volumetric water content values ranged from `r pn(vwcbegin$mean, 2)` ± `r pn(vwcbegin$sd, 2)` (min `r pn(vwcbegin$min, 2)`, max `r pn(vwcbegin$max, 2)`) at the beginning of the experiment to `r pn(vwcend$mean, 2)` ± `r pn(vwcend$sd, 2)` (min `r pn(vwcend$min, 2)`, max `r pn(vwcend$max, 2)`) at the end. Water filled pore space, assuming a particle density of `r PARTICLE_DENSITY` g cm^-3^, was `r pn(min(fluxdata$WFPS_percent), 0)`-`r pn(max(fluxdata$WFPS_percent), 0)`%.

After removal of known problematic data, and exclusion of statistical outliers, CO~2~ flux rates measured during the incubation ranged from `r pn(fluxsummary["flux_µgC_gC_day_min", "CO2"], 1)` µg C g C^-1^ day^-1^ (`r pn(fluxsummary["flux_µgC_g_day_min", "CO2"], 1)` µg C g soil^-1^ day^-1^) to a maximum of `r pn(fluxsummary["flux_µgC_gC_day_max", "CO2"], 1)` (`r pn(fluxsummary["flux_µgC_g_day_max", "CO2"], 2)`), with a mean of `r pn(fluxsummary["flux_µgC_gC_day_mean", "CO2"], 1)` (`r pn(fluxsummary["flux_µgC_g_day_mean", "CO2"], 1)`) over the 100 days. CH~4~ rates ranged from `r pn(fluxsummary["flux_µgC_gC_day_min", "CH4"]*1000, 2)` ng C g C^-1^ day^-1^ (`r pn(fluxsummary["flux_µgC_g_day_min", "CH4"]*1000, 2)` ng C g soil^-1^ day^-1^) to a maximum of `r pn(fluxsummary["flux_µgC_gC_day_max", "CH4"]*1000, 2)` (`r pn(fluxsummary["flux_µgC_g_day_max", "CH4"]*1000, 3)`), with a mean of `r pn(fluxsummary["flux_µgC_gC_day_mean", "CH4"]*1000, 2)` (`r pn(fluxsummary["flux_µgC_g_day_mean", "CH4"]*1000, 2)`).

These means conceal considerable variablility over the course of the incubation (**Figures 2 and 3**). In the linear mixed-effects model, CO~2~ was strongly influenced by incubation chamber temperature, core gravimetric water content, and percent soil N (all P < 0.05, and the latter two P < 0.001; **Table 1**). (Note that percent C and percent N were highly correlated (r ≥ 0.98) for these cores. Because percent N was a slightly stronger predictor, it was retained in the model while percent C was excluded.) The interaction between water content and percent N was also highly significant (`r pclean(summary(step_co2_lme)$tTable["WC_gravimetric:N_percent", "p-value"])`), with high-N cores having little relationship between water content and CO~2~ flux (data not shown).

@@

The cumulative production of C from CO~2~ was over `r minmag` order of magnitudes higher than that from CH~4~, with CO~2~:CH~4~ C ratios ranging from `r minratio` million in the `r mintemp` °C "`r maxtrt`" treatment, to `r maxratio` million in the `r maxtemp` °C "`r maxtrt`" treatment. Cumulative CO~2~ evolved was highly affected by temperature (`r pclean(co2_hsd$Temperature["p adj"])`), and "field moisture" cores emitted significantly more CO~2~ than the other two treatments at both temperatures (P < 0.001 for both). There was no difference between the 20 °C "drought" and "controlled drought" treatments (`r pclean(co2_hsd$Treatment["Drought-Controlled drought", "p adj"])`). "Drought" cores' cumulative production was `r pn(drought_effect["4", "Reduction"], 0)`% (4 °C) and `r pn(drought_effect["20", "Reduction"], 0)`% (20 °C) lower than the cores kept at field moisture. Neither temperature (`r pclean(ch4_hsd$Temperature[,"p adj"])`) nor treatment (mean `r pclean(mean(ch4_hsd$Treatment[,"p adj"]))`) was a significant factor in predicting cumulative CH~4~ fluxes.

>>>>

---
title: Temperature and moisture effects on greenhouse gas emissions from deep active-layer boreal soils
author: "Ben Bond-Lamberty^1^, A. Peyton Smith^2^, Vanessa Bailey^2^"
output: html_document
---
^1^Joint Global Change Research Institute, DOE Pacific Northwest National Laboratory, College Park, MD USA

@@

*Correspondence to*: Ben Bond-Lamberty (bondlamberty@pnnl.gov)

@@

Rapid climatic changes, rising air temperatures, and increased fires are expected to drive permafrost degradation and alter soil carbon (C) cycling in many high-latitude ecosystems. How these soils will respond to changes in their temperature, moisture, and overlying vegetation is highly uncertain, but critical to understand given the large soil C stocks in these regions. We used a laboratory experiment to examine how temperature and moisture control CO~2~ and CH~4~ emissions from mineral soils sampled from the bottom of the annual active layer, i.e. directly above permafrost, in an Alaskan boreal forest. Gas emissions from thirty cores, subjected to two temperatures and either field moisture conditions or experimental drought, were tracked over a 100-day incubation; we also measured a variety of physical and chemical characteristics of the cores. Gravimetric water content was `r pn(gwcbegin$mean, 2)` ± `r pn(gwcbegin$sd, 2)` (unitless) at the beginning of the incubation; cores at field moisture were unchanged at the end, but drought cores had declined to `r pn(gwcendmean["Drought"], 2)` ±  `r pn(gwcendsd["Drought"], 2)`. Carbon dioxide fluxes were strongly influenced by incubation chamber temperature, core water content, and percent soil nitrogen. Methane fluxes were most strongly correlated with percent nitrogen, but neither temperature nor water content was a significant first-order predictor of CH~4~ fluxes. The cumulative production of C from CO~2~ was over `r minmag` orders of magnitudes higher than that from CH~4~. The CO~2~ temperature sensitivity (i.e. Q~10~) was `r pn(Q10_cumulative["Field moisture", "CO2"], 1)` and `r pn(Q10_cumulative["Drought", "CO2"], 1)` for the field moisture and drought treatments, respectively. These results suggest that deep active-layer soils may be much more sensitive to changes in moisture than to temperature, a critical factor as discontinuous permafrost melts in interior Alaska. Deep but unfrozen high-latitude soils have been shown to be strongly affected by long-term experimental warming, and these results help us understand the future dynamics, and feedback potential, of such soils.

#### 1 Introduction

High latitude ecosystems are being subjected to rapid changes in climate {IPCC, 2013 #3916} and increases in fire frequency and intensity {Kasischke, 2010 #2631}, notably in northwestern North America and Alaska {Hinzman, 2005 #4315}{Ju, 2016 #4320}. This will have a wide variety of ecosystem effects {Alexander, 2016 #4414}: in particular, rising temperatures and increasing fire will likely result in permafrost degradation {Pastick, 2015 #4311}{Zhang, 2015 #4372}{Genet, 2013 #4412}{Helbig, 2016 #4413} and changes in soil temperature, with subsequent hydrology changes that will influence soil greenhouse gas (GHG) fluxes to the atmosphere. Such fluxes are a large component of the global C cycle and, because of the high C stocks of northern soils {Tarnocai, 2009 #2231}, could result in a significant and positive climate feedback {Treat, 2015 #4217}{Koven, 2011 #3092}{Schaefer, 2014 #4117}.

The magnitude, timing, and form-in particular as methane (CH~4~) or carbon dioxide (CO~2~)-of such any such feedback remain highly uncertain {Schuur, 2015 #4319}. While northern soils hold enormous quantities {Tarnocai, 2009 #2231} of potentially mineralizable soil organic carbon (SOC), vegetation and succession dynamics (for example, thermal insulation by mosses) promote permafrost resilience to even large temperature changes {Jorgenson, 2010 #2627}{Turetsky, 2012 #3405}. Such dynamics may however be disrupted by increased fire disturbance, particularly with more-intense fires {Johnstone, 2010 #2465}{Genet, 2013 #4412}. In addition, the stability of SOC is itself highly uncertain, as it depends on soil temperature and moisture, the ages of and ratio between the carbon (C) and nitrogen (N) pools {Weiss, 2015 #4317}, and its protection from competent microorganisms, enzymes, and resources {Bailey, 2012 #3624}, whether by organomineral sorption, chemical lability, or physical location {Schmidt, 2011 #3135}.

Temperature and moisture typically have strong and often interactive influences on soil GHG emissions. Laboratory incubations, field observations, and meta-analyses have documented increased fluxes of CO~2~, and under some conditions of CH~4~ {Olefeldt, 2013 #3456}, with rising temperature {Davidson, 2006 #1669}{Hashimoto, 2015 #4231}{Treat, 2015 #4217}. Greenhouse gas (GHG) responses to wetting and thawing dynamics are much less certain, with substantial variability between studies {Kim, 2012 #2863}. The anaerobic conditions common following permafrost thaw are expected to lower CO~2~ emissions but increase those of CH~4~ {Treat, 2015 #4217;Treat, 2014 #4213}, and such interactions are critical to examine in the course of long-term incubation experiments {Elberling, 2013 #3720}. More uniformly across the landscape, however, decadal warming, and perhaps drying, trends in Alaska {Bieniek, 2014 #4389} may counteract these effects. A critical question, then, is how the structure, chemistry, and microbial communities of current active-layer soils will respond to almost-certain temperature and moisture changes in the future {Xue, 2016 #4361}.

The goal of this study was to examine how temperature and moisture control GHG (CO~2~ and CH~4~) emissions from soils sampled from the bottom of the annual active layer–i.e., directly above permafrost–in an Alaskan boreal forest. Most previous studies have focused on surface soils or permafrost soils, neglecting deep active-layer soils that were identified as subject to strong effects from a two-decade warming experiment in the Alaskan Arctic {Sistla, 2013 #3666}. We also aimed to characterize the chemical and structural properties of these soils following a 100-day incubation at different temperatures, subjecting some cores to drying treatments. We hypothesized that (i) CO~2~ would be the dominant pathway for C loss in these largely aerobic soils; (ii) soils maintained at field moisture and high temperature would lose more C-CO~2~ than cores incubated at 4˚C, due to increased aerobic and anaerobic microbial activity; and (iii) core CH~4~ fluxes would be sensitive only to temperature, as no anaerobic conditions were imposed on the cores.

#### 2 Methods

**2.1 Field sampling**

The field component of this research took place in Caribou-Poker Creeks Research Watershed (CPCRW), part of the Bonanza Creek LTER (http://www.lter.uaf.edu/research/study-sites-cpcrw). CPCRW is located in the Yukon-Tanana Uplands northeast of Fairbanks, AK, a part of the boreal forest that has seen strong increases in air temperature and forest browning {Ju, 2016 #4320} over several decades. Annual average air temperature was -2.5 °C, and annual average precipitation 400 mm {Petrone, 2006 #4313}. The watershed's lowlands and north-facing slopes are dominated by black spruce (*Picea mariana* (Mill.) BSP), feathermoss (*Pleurozium schreberi* and others), and *Sphagnum* spp.; the drier south slopes tend to be deciduous with a mixture of trembling aspen (*Populus tremuloides* Michx.), paper birch (*Betula neoalaskana*), and patches of alder (*Alnus crispa*).

@@

**2.2 Laboratory incubation**

In the lab, the soil cores were stored at 4 °C for several days until they were weighed and prepared for incubation. At that point (11-12 August 2015), three fragmented or otherwise damaged cores were discarded, and the remaining cores were randomly assigned to one of six groups (N=6 in each group). These included two incubation temperatures of 4 and 20 °C, following the protocol of a number of previous boreal incubation studies {cf. Table 1 in \Treat, 2015 #4217}. Within each temperature there were two moisture treatments: one in which soil moisture was maintained at field conditions (~28% moisture by volume), and a drought treatment in which no water was added and cores were allowed to dry down to ~5% moisture by volume. The fifth group was a 20 °C "controlled drought" one, in which water was added so that these cores' moisture status would close match those of the 4 °C "drought" cores, which we anticipated would dry more slowly than their 20 °C counterparts. The final 6-core group was used for destructive, pre-incubation measurements including moisture content, pH, soil carbon and nitrogen, and bulk density. Subsamples were collected and stored at -20 °C for dissolved organic carbon measurements or air-dried for soil C and N (see below).

On 18 August 2015 cores were placed into one of two growth chambers (Conviron Control Systems BDW80, Winnipeg, Canada) maintained at 4 and 20 °C temperatures and 70% relative humidity and allowed to equilibrate for two weeks. Starting on 31 August 2015 we measured the cores' mass and GHG (CH~4~ and CO~2~) emissions four times in the first week, then twice per week for the first month, and then once per week for the rest of the 100-day incubation. Throughout the incubation, cores with a 200 µm mesh screen fit to the base were mounted on porous ceramic plates (Soil Moisture Equipment Corp., Santa Barbara, CA, USA) so that, when the plates were placed in contact with water, water would move up into the cores via capillary action. The "drought" cores were mounted on dry plates, but not allowed to drop below 5% water content. When necessary, cores received additional wetting from the top to maintain their water status at the desired level.

@@

The incubation experiment concluded on 9 December 2015, following the final CO~2~ and CH~4~ readings. Each soil core was maintained at the treatment-dependent temperature and moisture content (by mass) until removed for destructive sampling, December 14-18, 2015. Sub-samples were collected and composited throughout each soil core for dissolved organic carbon analysis (110 ± 24 g dry mass equivalent) and dry-mass calculations (~28 g each). The remaining core material was air-dried and separated into particles (>2 mm diameter) and soil (≤2 mm) using a U.S. Standard Test Sieve No. 10 (Fisherbrand,  Pittsburg, PA, USA). The dry mass and volume of soil were used in calculations of gravimetric and volumetric soil moisture content, respectively {Gardner, 1986 #4345}. Soil volume was calculated as the total core volume minus the volume of particles >2 mm diameter, with the latter determined by water displacement. Air-dried soil and sub-samples stored at -20 °C were sent to the Agricultural and Environmental Services Laboratory at the University of Georgia Extension in February 2016 for total C, N, and dissolved organic C. Samples were combusted in an oxygen atmosphere at 1350 °C, and measured for gaseous C and N using a Elementar Vario Max CNS. DOC was measured using a Shimadzu 5000 TOC Analyzer.

**2.3 Data and statistical analysis**

@@

@@

Anomalous data were excluded based on their gas fluxes being more than `r CO2_EXCLUDE_DEVS` (for CO~2~) or `r CH4_EXCLUDE_DEVS` (for CH~4~) mean absolute deviations {Davies, 1993 #3138} from the treatment mean within a 10-day period, for a given treatment and temperature. We excluded `r mad_outlier_count` of `r pre_outlier_count` (`r pn(mad_outlier_count/pre_outlier_count*100, 1)`%) measurements for this reason. If the coefficient of variability (CV) of fluxes from any core on a single day exceeded `r pn(MAX_CO2_CV*100, 0)`%, a value chosen based on the distribution of CVs across all cores, the entire core was excluded for that day (`r cv_outlier_count` data points, `r pn(cv_outlier_count/pre_outlier_count*100, 1)`%). Other data (`r pn(rn/sdn*100, 1)`%) were removed because of known instrument problems, e.g. the analyzer was left running after leaving a chamber. The final number of valid flux samples from the 100-day incubation was `r sum(!fluxdata$outlier)`.

The effects of temperature, gravimetric water content, percent C, percent N, and DOC concentration on instantaneous gas fluxes were evaluated using a linear mixed-effects model fit by the R function *lme* in the R 'nlme' package, version `r packageVersion("nlme")`. Because the dependent variable (CO~2~ or CH~4~ flux) was non-normally distributed, it was transformed using a natural-logarithm (+`r FLUX_ADDITION` µg C g C^-1^ day^-1^ to ensure all positive fluxes, following Treat et al. 2015) transformation. Soil core was treated as a random effect in the model. We then used performed stepwise model selection by Akaike's information criterion (AIC) using the *stepAIC* function in the R 'MASS' package, version `r packageVersion("MASS")`. A linear mixed-effects model was also used to evaluate the effect of treatment on core water content.

@@

#### 3 Results

```{r coredata, echo=FALSE}
# Summarize a few bulk density metrics
bd <- with(coredata, SoilDryMass_g / SoilVolume_cm3)
schist <- with(coredata, NonsoilMassLarge_g / (NonsoilMassLarge_g + SoilDryMass_g)) * 100
sdm <- coredata$SoilDryMass_g
```

The `r nrow(coredata)` experimental cores had a bulk density of `r pn(mean(bd), 2)` ± `r pn(sd(bd), 2)` (mean ± sd) g cm^-3^. Large (>2 mm) particles, primarily schist, comprised `r round(mean(schist), 0)`% ± `r round(sd(schist), 0)`% of the cores' total mass. Soil (≤2 mm) dry mass was `r round(mean(sdm), 0)` ± `r round(sd(sdm), 0)` g. Sample DOC was `r pn(cndata_summary["DOC_mg_kg_mean"], 2)` ± `r pn(cndata_summary["DOC_mg_kg_sd"], 2)` mg kg^-1^. Carbon content was `r pn(cndata_summary["C_percent_mean"], 2)`% ± `r pn(cndata_summary["C_percent_sd"], 2)`%, while N content was `r pn(cndata_summary["N_percent_mean"], 2)`% ± `r pn(cndata_summary["N_percent_sd"], 2)`%. Neither temperature nor moisture treatment exerted any significant effect on these highly variable properties (P > 0.1 for all).

Gravimetric water content was `r pn(gwcbegin$mean, 2)` ± `r pn(gwcbegin$sd, 2)` (min `r pn(gwcbegin$min, 2)`, max `r pn(gwcbegin$max, 2)`) at the beginning of the incubation (**Figure 1**). "Field moisture" cores were on average unchanged (`r pn(gwcendmean["Field moisture"], 2)` ± `r pn(gwcendsd["Field moisture"], 2)`) at the end of the incubation, but both the drought treatments, which did not differ from each other in their effect on gravimetric water content (P = `r pn(p_trt_wc, 3)`), had declined to `r pn(gwcendmean["Drought"], 2)` ±  `r pn(gwcendsd["Drought"], 2)`. Volumetric water content values ranged from `r pn(vwcbegin$mean, 2)` ± `r pn(vwcbegin$sd, 2)` (min `r pn(vwcbegin$min, 2)`, max `r pn(vwcbegin$max, 2)`) at the beginning of the experiment to `r pn(vwcend$mean, 2)` ± `r pn(vwcend$sd, 2)` (min `r pn(vwcend$min, 2)`, max `r pn(vwcend$max, 2)`) at the end across all cores. Water filled pore space, assuming a particle density of `r PARTICLE_DENSITY` g cm^-3^, was `r pn(min(fluxdata$WFPS_percent), 0)`-`r pn(max(fluxdata$WFPS_percent), 0)`%.

Carbon dioxide fluxes during the incubation ranged from `r pn(fluxsummary["flux_µgC_gC_day_min", "CO2"], 1)` µg C g C^-1^ day^-1^ (`r pn(fluxsummary["flux_µgC_g_day_min", "CO2"], 1)` µg C g soil^-1^ day^-1^) to a maximum of `r pn(fluxsummary["flux_µgC_gC_day_max", "CO2"], 1)` (`r pn(fluxsummary["flux_µgC_g_day_max", "CO2"], 2)`), with a mean of `r pn(fluxsummary["flux_µgC_gC_day_mean", "CO2"], 1)` (`r pn(fluxsummary["flux_µgC_g_day_mean", "CO2"], 1)`) over the 100 days. CH~4~ rates ranged from `r pn(fluxsummary["flux_µgC_gC_day_min", "CH4"]*1000, 2)` ng C g C^-1^ day^-1^ (`r pn(fluxsummary["flux_µgC_g_day_min", "CH4"]*1000, 2)` ng C g soil^-1^ day^-1^) to a maximum of `r pn(fluxsummary["flux_µgC_gC_day_max", "CH4"]*1000, 2)` (`r pn(fluxsummary["flux_µgC_g_day_max", "CH4"]*1000, 3)`), with a mean of `r pn(fluxsummary["flux_µgC_gC_day_mean", "CH4"]*1000, 2)` (`r pn(fluxsummary["flux_µgC_g_day_mean", "CH4"]*1000, 2)`).

These means conceal considerable variability over the course of the incubation (**Figures 2 and 3**). In the linear mixed-effects model, CO~2~ was strongly influenced by incubation chamber temperature, core gravimetric water content, and percent soil N (all P < 0.05, and the latter two P < 0.001; **Table 1**). (Note that percent C and percent N were highly correlated (r ≥ 0.98) for these cores. Because percent N was a slightly stronger predictor, it was retained in the model while percent C was excluded.) The interaction between water content and percent N was also highly significant (`r pclean(summary(step_co2_lme)$tTable["WC_gravimetric:N_percent", "p-value"])`), with high-N cores having little relationship between water content and CO~2~ flux (data not shown).

@@

The cumulative production of C from CO~2~ (**Figure 4**) was over `r minmag` order of magnitudes higher than that from CH~4~, with CO~2~:CH~4~ C ratios ranging from `r minratio` million in the `r mintemp` °C "`r maxtrt`" treatment, to `r maxratio` million in the `r maxtemp` °C "`r maxtrt`" treatment. Cumulative CO~2~ evolved was highly affected by temperature (`r pclean(co2_hsd$Temperature["p adj"])`), and "field moisture" cores emitted significantly more CO~2~ than the other two moisture treatments at both temperatures (P < 0.001 for both, with no significant interactive effect). There was no difference between fluxes from the 20 °C "drought" and "controlled drought" treatments (`r pclean(co2_hsd$Treatment["Drought-Controlled drought", "p adj"])`). "Drought" cores' cumulative production was `r pn(drought_effect["4", "Reduction"], 0)`% (4 °C) and `r pn(drought_effect["20", "Reduction"], 0)`% (20 °C) lower than the cores kept at field moisture. Neither temperature (`r pclean(ch4_hsd$Temperature[,"p adj"])`) nor moisture treatment (mean `r pclean(mean(ch4_hsd$Treatment[,"p adj"]))`) was a significant factor in predicting cumulative CH~4~ fluxes.

The cumulative flux numbers above result in CO~2~ temperature sensitivity (Q~10~) values of `r pn(Q10_cumulative["Field moisture", "CO2"], 1)` and `r pn(Q10_cumulative["Drought", "CO2"], 1)` for the field moisture and drought treatments, respectively; the corresponding Q~10~ values  based on cumulative CH~4~ were `r pn(Q10_cumulative["Field moisture", "CH4"], 1)` and `r pn(Q10_cumulative["Drought", "CH4"], 1)`. Computing Q~10~ values based on fluxes normalized by water filled pore space changed these values only slightly: to `r pn(wfpsQ10_cumulative["Field moisture", "CO2"], 1)` and `r pn(wfpsQ10_cumulative["Drought", "CO2"], 1)` for CO~2~, respectively, and `r pn(wfpsQ10_cumulative["Field moisture", "CH4"], 1)` and `r pn(wfpsQ10_cumulative["Drought", "CH4"], 1)` for CH~4~.

#### 4 Discussion
